---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
author: "Matthias Gondan and Irene Alfarone
         (Department of Psychology, Universität Innsbruck, Austria)"
date: "2023-01-09"
output: rmarkdown::html_vignette
bibliography: RJreferences.bib
preamble:
- \usepackage{longtable}
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# To do
#
# paste(x, y, z)
# Spacing
# Comma separated lists
# Ellipsis
# Radicals
# Subsets
# Greek letters in MathJax
# Typeface
# Stylechanges
# Phantoms

library(mathml)

mathout <- function(..., flags=NULL, env=parent.frame())
{
  if(knitr::is_html_output())
    return(mathml(..., flags=c(flags, list(cat=TRUE)), env=env))

  if(knitr::is_latex_output())
    return(mathjax(..., flags=c(flags, list(cat=TRUE)), env=env))

  warning("mathout: no knitr output specified")  
  mathjax(..., flags=c(flags, list(cat=TRUE)), env=env)
}

catmath <- function(term, ...)
{
  s <- substitute(term)
  mathout(canonical(s), ...)
}
```

This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
thereby allowing for a pretty mathematical representation of R\ objects and
functions in data analyses, scientific reports and interactive web content. In
the RMarkdown document rendering language, R\ code and mathematical content
already exist side-by-side. The present package allows to use the same
R\ objects for both data analysis and typesetting in documents or web content.
This tightens the link between the statistical analysis and its verbal
description or symbolic representation, which is another step towards
reproducible science. User-defined hooks allow to extend the package by
mapping specific variables or functions to new MathML and LaTeX entities. A
few working examples illustrate the use of the package in a scientific report
and in on an interactive web page.

# Introduction

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the mapping of built-in language elements to their
mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
such as \(P_{\mathrm{Bi}}(X \le k; N, p)\) require explicit Latex commands, that
is, `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for minimalistic use
cases, these commands are tedious to type in and their source code is hard to
read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and reduces mistakes, as will be illustrated below.

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \pkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
studies for the use of the package in a scientific report and 
a **ShinyApp web page**. Section\ 6 concludes with a discussion and ideas for
further development. A complete list of the R\ elements covered by the package
is found in the Appendix.

*shinyapp missing so far

# Background

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R\ language. The \pkg{rolog} package [@rolog] bridges this
gap by providing an interface to a SWI-Prolog distribution in R. The
communication between the two systems is mainly in the form of queries from
R\ to Prolog, but two predicates allow Prolog to ring back and evaluate terms in
R.

For a first illustration of the \pkg{mathml} package, we consider the binomial
probability.

```{r}
term <- quote(pbinom(k, N, p))
term
```

The term is quoted to avoid its immediate evaluation (which would raise an error
anyway since the variables `k`, `N`, `p` have not yet been defined). Experienced
readers will remember that the quoted expression above is a short form for

```r
term <- call("pbinom", as.name("k"), as.name("N"), as.name("p"))
```

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
can eventually be evaluated with `eval()`,

```{r}
k <- 10
N <- 22
p <- 0.4
eval(term)
```

The R\ package \pkg{mathml} can now be used to render the call in MathML, that
is the dialect for mathematical elements on HTML webpages or in MathJax/LaTeX,
as shown below (some of the curly braces are not really needed in this simple
example, but are necessary in edge cases).

```{r}
library(mathml)
mathjax(term)
```

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\ code chunk, as is done in the next example. The
R\ function `mathout()` is a wrapper that invokes `mathml()` for HTML output
and `mathjax()` for LaTeX output.

```{r, results='asis'}
mathout(term)
```

At the Prolog end, a predicate `math/2` translates the call `pbinom(K, N, Pi)`
into a "function" `fn/2` with the name `P_Bi`, one argument `X =< K`, and the
two parameters `N` and `Pi`.

```prolog
math(pbinom(K, N, Pi), M)
 => M = fn(subscript('P', "Bi"), (['X' =< K] ; [N, Pi])).
```

Thus, the predicate `math/2` could be considered a "macro" that translates a
mathematical element (here, `pbinom(K, N, Pi)`) to a different mathematical
element, namely `fn(Name, (Args ; Pars))`. The low-level predicate `ml/3` is
used to convert these basic elements to MathML.

```prolog
ml(Flags, fn(Name, (Args ; Pars)), M)
 => ml(Flags, Name, N),
    ml(Flags, paren(list(op(;), [list(op(','), Args), list(op(','), Pars)])), X),
    M = mrow([N, mo(&(af)), X]).
```

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters, enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the errors below).

# Package \pkg{mathml} in practice

\pkg{mathml} is an R\ package for pretty mathematical representation of
R\ functions and objects in data analysis, scientific reports and interactive
web content. The currently supported features are listed below, roughly
following the order proposed by [@murrell2000].

## Basic elements

\pkg{mathml} handles the basic elements of everyday mathematical expressions,
such as numbers, Latin and Greek letters, multi-letter identifiers, accents,
subscripts, and superscripts. 
 
```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + hat(b)[i, j]^2L)
mathout(term)

term <- quote(NaN + NA + TRUE + FALSE + Inf + (-Inf))
mathout(term)
```

An expression such as `1 + -2` may be considered unsatisfactory from an
aesthetical perspective. It is correct R\ syntax, though, and is reproduced
accordingly, without the parentheses. Parentheses around negated numbers or
symbols can be added as shown for `-Inf`.

Note that an R function `hat()` does not exist in base R, it is provided by the
package for convenience and points to the identity function.

## Operators and parentheses

Arithmetic operators and parentheses are translated as they are, as illustrated
below.

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
mathout(term)

term <- quote(dot(a, b) + frac(1L, nodot(c, d + e)) + dfrac(1L, times(g, h)))
mathout(term)
```

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines alternative multiplication
functions `dot()`, `nodot()`, and `times()` that calculate the product and
produce the respective multiplication signs. Similarly, `frac()` and `dfrac()`
can be used for small and large fractions.

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unecessary
parentheses around `l + m` above. Thes parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

```{r}
term <- quote(a^(b + c))
paste(term)
```

For the present purpose, this feature is unfortunate because extra parentheses
around `b + c` are not needed. The preferred result is obtained by using the
functional form `quote('^'(k, l + m))` of the power, or curly braces as a
workaround (see `p + q` above).

## Custom operators

Whereas in standard infix operators, the parentheses typically follow the rules
for precedence, undesirable results may be obtained in custom operators.

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N)))
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. As in the previous example, the solution is to
either use the functional form of the operator or a curly brace to enforce the
correct operator precedence.

More operators are shown in Table \@ref(tab:custom-operators), including the
suggestions by Murrell and Ihaka [-@murrell2000] for graphical annotations and
arrows in R\ figures.

```{r custom-operators, echo=FALSE}
op1 <- list(
  cross=quote(A %*% B),
  dot=quote(A %.% B),
  kronecker=quote(A %x% B),
  'integer division'=quote(A %/% B),
  modulo=quote(A %% B),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(A == B),
  assign=quote(A <- B))

m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

op1 <- as.character(op1)
if(knitr::is_latex_output())
  op1 <- sapply(op1, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op1 <- sapply(op1, FUN=knitr:::escape_html)

op2 <- list(
  different=quote(A != B),
  tilde=quote(A ~ B),
  approx=quote(A %~~% B),
  equiv=quote(A %==% B),
  congr=quote(A %=~% B),
  prop=quote(A %prop% B),
  'in'=quote(A %in% B),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  crossprod=quote(crossprod(A, B)),
  intersect=quote(is.null(A)))

m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

op2 <- as.character(op2)
if(knitr::is_latex_output())
  op2 <- sapply(op2, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op2 <- sapply(op2, FUN=knitr:::escape_html)

op3 <- list(
  twosidearrow=quote(A %<->% B),
  rightarrow=quote(A %->% B),
  leftarrow=quote(A %<-% B),
  uparrow=quote(A %up% B),
  down=quote(A %down% B),
  iff=quote(A %<=>% B),
  rArr=quote(A %=>% B),
  lArr=quote(A %<=% B),
  dblup=quote(A %dblup% B),
  dbldown=quote(A %dbldown% B))

m3 <- lapply(op3, FUN=mathout, flags=list(cat=FALSE))

op3 <- as.character(op3)
if(knitr::is_latex_output())
  op3 <- sapply(op3, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op3 <- sapply(op3, FUN=knitr:::escape_html)

t <- cbind(Operator=op1, Output=m1,
  Operator=op2, Output=m2,
  Operator=op3, Arrow=m3)

knitr::kable(t, caption="Custom operators in mathml",
  row.names=FALSE, escape=FALSE)
```

## Builtin functions

There is support for most functions from \pkg{base}, with adequate use and omission
of parentheses.

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
mathout(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + ceiling(x))
mathout(term)
```

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from \pkg{stats}. 

```{r base-stats, echo=FALSE}
op1 <- list(
  sin=quote(sin(x)),
  cosh=quote(cosh(x)),
  tanpi=quote(tanpi(alpha)),
  asinh=quote(asinh(alpha)),
  log=quote(log(p)),
  log1p=quote(log1p(x)),
  logb=quote(logb(x, e)),
  exp=quote(exp(x)),
  expm1=quote(expm1(x)),
  choose=quote(choose(n,k)),
  lchoose=quote(lchoose(n, k)),
  factorial=quote(factorial(n)), 
  lfactorial=quote(lfactorial(n)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(X)),
  absolute=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

op1 <- as.character(op1)
if(knitr::is_latex_output())
  op1 <- sapply(op1, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op1 <- sapply(op1, FUN=knitr:::escape_html)

op2 <- list(
  dbinom=quote(dbinom(k, N, pi)),
  pbinom=quote(pbinom(k, N, pi)),
  qbinom=quote(qbinom(p, N, pi)),
  dpois=quote(dpois(k, lambda)),
  ppois=quote(ppois(k, lambda)),
  qpois=quote(qpois(p, lambda)),
  dexp=quote(dexp(x, lambda)),
  pexp=quote(pexp(x, lambda)),
  qexp=quote(qexp(p, lambda)),
  dnorm=quote(dnorm(x, mu, sigma)),
  pnorm=quote(pnorm(x, mu, sigma)),
  qnorm=quote(qnorm(alpha/2L)),
  pchisq=quote(1L-pchisq(x, 1L)),
  qchisq=quote(qchisq(1L-alpha, 1L)),
  pt=quote(pt(t, N-1L)),
  qt=quote(qt(alpha/2L, N-1L)))
m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

op2 <- as.character(op2)
if(knitr::is_latex_output())
  op2 <- sapply(op2, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op2 <- sapply(op2, FUN=knitr:::escape_html)

t <- cbind(Function=op1, Output=m1, Function=op2, Output=m2)
knitr::kable(t, caption="R functions from base and stats",
  row.names=FALSE, escape=FALSE)
```

## Custom functions

For self-written functions, matters are a bit more complicated. The name of a
self-written function is not transparent in R, since only the function body is
represented, as illustrated by the redefinition of `sign` below. The function is
often a nested R "call" of the form `'{'(L)`, with `L` being a list of
commands. \pkg{mathml} provides very limited support for typical control
structures.

```{r, results='asis'}
sgn <- function(x)
{
  if(x == 0L) return(0L)
  if(x < 0L) return(-1L)
  if(x > 0L) return(1L)
}

term <- quote(mysign <- sgn(x))
mathout(term)

hook(quote(mysign), quote(sg(x)))
term <- call("<-", quote(mysign), sgn)
mathout(term)
```

The hook (see below) is used to display \(\mathrm{sg}(x) = \{\ldots\) instead
of just \(s = \{\).

## Indices and powers

Indices in square brackets are rendered as subscripts, powers are rendered as
superscript. Moreover, \pkg{mathml} defines the
functions `subscript()`, `superscript()` and `subsuperscript()` that simply
return their first argument. The other arguments serve as decorations, for
example, for summation and product signs.

```{r, results='asis'}
term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
mathout(term)

term <- quote(log(subsupscript(prod(L[i]), i==1L, N)) <- 
          subsupscript(sum(log(L[i])), i==1L, N))
mathout(term)
```

## Ringing back to R

R\'s `integrate` function takes a number of arguments, the most important ones
being the function to integrate, and the lower and the upper bound of the
integration.

```{r, results='asis'}
term <- quote(integrate(sin, 0L, 2L*pi))
mathout(term)
eval(term)
```

For mathematical typesetting in the form
of \(\int f(x)\, dx\), \pkg{mathml} needs to find out the name of the
integration variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R\ from Prolog. In the example above, this
predicate evaluates `formalArgs(args(sin))`, which returns the names of the
arguments of `sin`, namely, `x`.

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. R's `integrate()` can handle both. However, as already mentioned in
the previous subsection, \pkg{mathml} would be unable to find out the function
name, it needs the symbol.

## Names and order of arguments

One of R's great features is the possibility to refer to function arguments by
their names, not only by their position in the list of arguments. At the other
end, most Prolog-powered handlers for the R\ calls are rather rigid, for
example, `integrate/3` accepts exactly three arguments in a particular order and
without names, that is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print
the desired result.

To "canonicalize" function calls with named arguments and arguments in unusual
order, \pkg{mathml} provides an auxilliary R\ function `canonical(f, drop)` that
reorders the argument list of calls to known R\ functions and, 
if `drop=TRUE` (which is the default), also removes the names of the arguments.

```{r}
term <- quote(integrate(lower=0L, upper=2L*pi, sin))
canonical(term)
```

```{r, results='asis'}
mathout(canonical(term))
```

This function can be used to feed mixtures of partially named and positional
arguments into the renderer. For details, see the R\ function `match.call()`.

## Matrices and Vectors

Of course, \pkg{mathml} also supports matrices and vectors.

```{r, results='asis'}
v <- 1:3
mathout(call("t", v))

A <- matrix(data=11:16, nrow=2, ncol=3)
B <- matrix(data=21:26, nrow=2, ncol=3)
term <- call("+", A, B)
mathout(term)
```

Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
the output---instead of the desired matrix representation.

## Abbreviations

We consider the \(t\)-statistic for independent samples with equal variance. To
avoid clutter in the equation, the pooled variance \(s^2_{\mathrm{pool}}\) is
abbreviated, and a comment is given with the expression
for \(s^2_{\mathrm{pool}}\). For this purpose, \pkg{mathml} provides a
function `denote(abbr, expr, info)`, with `expr` actually being
evaluated, `abbr` being rendered, plus a comment of the
form "with `expr` denoting `info`". The hooks are described in the next section.

```{r, results='asis'}
hook(quote(m_A), quote(mean(X)["A"])); hook(quote(s2_A), quote(s["A"]^2L));
hook(quote(n_A), quote(n["A"]))
hook(quote(m_B), quote(mean(X)["B"])); hook(quote(s2_B), quote(s["B"]^2L))
hook(quote(n_B), quote(n["B"])); hook(quote(s2_p), quote(s["pool"]^2L))

term <- quote(t <- dfrac(m_A - m_B, 
    sqrt(denote(s2_p, frac((n_A - 1L)*s2_A + (n_B - 1L)*s2_B, n_A + n_B - 2L),
                "the pooled variance.") * (frac(1L, n_A) + frac(1L, n_B)))))
mathout(term)
```

The term is evaluated below. `print()` is needed because the return value of the
assignment in the term `t <- dfrac(...)` is invisible.

```{r}
m_A <- 1.5; s2_A <- 2.4^2; n_A <- 27; m_B <- 3.9; s2_B <- 2.8^2; n_B <- 20
print(eval(term))
```

## Customizing the package

In typical R\ functions, variable names are typically longer than just single
letters, which may yield unsatisfactory results in the mathematical output.

```{r, results='asis'}
term <- quote(pbinom(successes, Ntotal, prob))
mathout(term)

hook(quote(successes), quote(k))
hook(quote(Ntotal), quote(N))
hook(quote(prob), quote(pi))
mathout(term)
```

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. Care should be taken to
avoid recursive hooks such as `hook(quote(s), quote(s["A"]))` that endlessly
replace the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

more stuff: math/2, ml/3, jax/3

# A case study {#case-study}

This case study describes a model by Schwarz [-@schwarz1994] from mathematical
Psychology using the features of package \pkg{mathml}. Schwarz presents a new
explanation of redundancy gains that occur when observers respond to stimuli of
different sources, and the same information is presented on two or more
channels. In Schwarz' model, decision-making builds on a process of noisy
accumulation of information over time [e.g., @ratcliff2016]. In redundant
stimuli, the model assumes a superposition of channel-specific diffusion
processes that eventually reach an absorbing barrier to elicit the response. For
a detailed description the reader may refer to the original article.

## The model 

```{r, echo=FALSE}
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (AB, redundant stimuli), with the redundant stimuli
being presented either simultaneously or with onset asynchrony \(\tau\). The
channel activation is described as a two-dimensional Wiener process with
drifts \(\mu_i\), variances \(\sigma^2_i\), and initial
conditions \(X_i(t = 0) = 0, i = \mathrm{A, B}\). The channels may be correlated
with \(\rho_{\mathrm{AB}}\), but we assume \(\rho_{\mathrm{AB}} = 0\) for
simplicity.

A response is elicited when the process reaches an absorbing barrier \(c > 0\)
for the first time. In single-target trials, the first passages at \(c\) are
expected at

```{r, results='asis'}
hook(call("ED_single", expression(c), expression(mu)), 
     call("E", call(";", quote(D), call(",", expression(c), expression(mu)))))

ED_single <- function(c, mu)
  dfrac(c, mu)

mathout(call("=", quote(ED_single(c, mu)), ED_single))
```

In redundant stimuli, the activation of the channel-specific diffusion processes
adds up, \(X_{\mathrm{AB}}(t) = X_{\mathrm A}(t) + X_{\mathrm B}(t)\), hence
the name, superposition. \(X_{\mathrm{AB}}(t)\) is again a Wiener process with
drift \(\mu_{\mathrm A} + \mu_{\mathrm B}\) and
variance \(\sigma^2_{\mathrm A} + \sigma^2_{\mathrm B}\). For the expected
first-passage time, we have

```{r, results='asis'}
mathout(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
```

For asynchronous stimuli, Schwarz [-@schwarz1994; Eq. 10] derived the expected
first-passage time as a function of the stimulus onset asyncrony \(\tau\),

```{r, results="asis"}
hook(call("ED_async", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)), 
     call("E", call(";", call("[", quote(D), expression(tau)), 
                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B)))))

ED_async <- function(tau, c, mu_A, sigma_A, mu_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) *
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

mathout(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
```

For negative onset asynchrony (i.e., B before A), the parameters need to be
switched.

```{r, results='asis'}
hook(call("ED", expression(tau), expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)), 
     call("E", call(";", call("[", quote(D), expression(tau)), 
                    call(",", expression(c), expression(mu_A), expression(sigma_A), expression(mu_B), expression(sigma_B)))))

ED <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{
  if(tau == Inf) return(ED_single(c, mu_A)) ;
  if(tau == -Inf) return(ED_single(c, mu_B)) ;
  if(tau == 0L) return(ED_single(c, mu_A + mu_B)) ;
  if(tau > 0L) return(ED_async(tau, c, mu_A, sigma_A, mu_B)) ;
  if(tau < 0L) return(ED_async(abs(tau), c, mu_B, sigma_B, mu_A))
}

mathout(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
```

The observable response time is assumed to be the sum of \(D\), the time
employed to reach the threshold for the decision, and a residual \(M\) denoting
other processes such a motor preparation and execution. Correspondingly, the
expected response time amounts to

```{r, results='asis'}
ET <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  ED(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M

mathout(call("=", quote(E(T[tau])), ET))
```

## Methods

Schwarz [-@schwarz1994] applied the model to data from a redundant signals task
with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where \(0\) refers to the synchronous condition, and \(\pm\infty\) to the
single-target presentations. Each condition was replicated 400 times. The
observed mean response times and their standard deviations are given in
Table \@ref(tab:miller-data) below.

```{r miller-data, echo=FALSE}
tau  <- list(-Inf, -167L, -133L, -100L, -67L, -33L, 0L, 33L, 67L, 100L, 133L, 167L, Inf)
m    <- c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
s    <- c( 56,  58,  40,  40,  32,  28,  28,  28,  26,  30,  32,  34,  92)
n    <- c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400)

ttau <- lapply(tau, FUN=mathout, flags=list(cat=FALSE))
tm   <- lapply(as.integer(m), FUN=mathout, flags=list(cat=FALSE))
ts   <- lapply(as.integer(s), FUN=mathout, flags=list(cat=FALSE))
tn   <- lapply(as.integer(n), FUN=mathout, flags=list(cat=FALSE))

t <- cbind('$\\tau$'=ttau, '$m$'=tm, '$s$'=ts, '$n$'=tn)
knitr::kable(t, caption="Miller data", row.names=FALSE, escape=FALSE)
```

Assuming that the model is correct, the observable mean reaction times follow an
approximate Normal distribution around the model prediction \(E(T)\). We can,
therefore, use a standard goodness-of-fit measure by \(z\)-standardisation.

```{r, results='asis'}
z <- function(m, s, n, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(m - denote(E, ET(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M),
              "the expected mean response time"),
    s / sqrt(n))

mathout(call("=", quote(z[tau]), z))
```

The overall goodness-of-fit is the sum of the squared \(z\)-statistics for each
onset asynchrony. Assuming again that the architecture of the model is correct,
but the parameters are adjusted to the data, it follows
a \(\chi^2(8~\mathrm{df})\)-distribution. The degrees of freedom result from
the 13 predicted mean response times, minus 5 free parameters.

$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\ df)$.

```{r, results='asis'}
gof <- function(par, tau, m, s, n)
{
  zn <- Vectorize(z, vectorize.args='tau')
  zi <- zn(m, s, n, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
  sum(zi^2L)
}

mathout(gof)
```

```{r, results="asis", echo=FALSE}
lsq = function(tau, M, s, N)
{
  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
  optim(start, fn = gof,
    tau = tau, m = M, s = s, n = N, hessian = TRUE, control = list(fnscale = 1))
}

mathout(lsq)

# Parameters' estimates
fit <-  lsq(tau, m, s, n)

est = fit$par
se = sqrt(diag(solve(fit$hessian)))
cilo = qnorm(0.025, est, se)
ciup = qnorm(0.975, est, se)

# Lower bound
#cilo

# Upper bound
#ciup

```

### Results

The calculated Goodness of Fit value is $X^2= 30.54$, $p= 0.0002$. Additionally, we proceeded with the parameters' estimation with the Least Squares Method.

The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz [-@schwarz1994].

Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i. $\tau= 100$) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model. Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.


## Errors

An additional feature that this package offers is related to its application in educational scenarios. The functions here reported are intended to identify and highlight possible errors that students make, as well as provide the correct answer for the task.

In this first example the function highlights that the student forgot to add the left-hand side of the expression and it does that by striking the missing part. In this case the letter A and the operator `+`.
Similarly the function `omit_right` does the same for the omissions in the right-hand side of the expression. Finally the function `omit()` can be used, for instance, to highlight the omission in the elements of a list.

```{r, results="asis"}
t <- quote(omit_left(a + b))
mathout(t)
```

```{r, results="asis"}
t <- quote(omit_right(a + b))
mathout(t)
```

```{r, results="asis"}
t <- list(quote(a), quote(b), quote(omit(c)))
mathout(t)
```

```{r omit, echo=FALSE}
op <- list(left=quote(omit_left(a + b)), right=quote(omit_right(a + b)), list=quote(omit(c)))
m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))

op <- as.character(op)
if(knitr::is_latex_output())
  op <- sapply(op, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op <- sapply(op, FUN=knitr:::escape_html)

t <- cbind(Operator=op, Output=m)
knitr::kable(t, caption="Omit", row.names=FALSE, escape=FALSE)
```

The functions `add_left()`, `add_right()`, `add()` reported below follow the same way of reasoning, but they account for the mistakes made in adding incorrect elements in the expression. They highlight the redundant element placing it in a box. 

```{r, results="asis"}
t <- quote(add_left(a + b))
mathout(t)
```

```{r, results="asis"}
t <- quote(add_right(a + b))
mathout(t)
```

```{r, results="asis"}
t <- list(quote(a), quote(b), quote(add(c)), quote(d))
mathout(t)
```

```{r add, echo=FALSE}
op <- list(left=quote(add_left(a + b)), right=quote(add_right(a + b)))
m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))

op <- as.character(op)
if(knitr::is_latex_output())
  op <- sapply(op, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op <- sapply(op, FUN=knitr:::escape_html)
t <- cbind(Operator=op, Output=m)

knitr::kable(t, caption="Add", row.names=FALSE, escape=FALSE)
```

Finally the function `instead( , )` place a curly bracket under the incorrect element  and provides its correct version under the curly bracket. In order to provide the correction with this function one must fill it with its wrong and correct terms in this order: `instead(wrong, correct)`. An example of its use its provided below.

```{r, results="asis"}
t <- quote(instead(a, b) + c)
mathout(t)
```

It is also possible to use flags for context-dependent rendering. The `highlight` flag is the default one. It shows where the error is and what is its solution. The `ignore` flag allows the program to ignore the fact that there is an error and print the expression as if it were correct. The third flag, `fix`, tells you what is the term that you are supposed to fix without giving the solution. Finally, the last flag `asis` returns the incorrect expression as the student wrote it, without providing any solutions or corrections.  Below is provided the code with its comment and an example of use.

Let us consider for example a student that instead of writing `b + c` wrote `a+c`. 

With the `asis` flag we report the incorrect expression provided by the student: 

```{r, results="asis"}
# It prints the incorrect version
t <- quote(instead(a, b) + c)
mathout(t, flags=quote(error(asis)))
```

Then, we may want to underline the incorrect term and provide the correct solution. 

```{r, results="asis"}
# It highlights what has been mistakenly inserted
mathout(t, flags=quote(error(highlight))) # default
```

Alternatively, we can put the correct solution in a box to let the student focus on its correction.

```{r, results="asis"}
# It highlights what has been corrected
mathout(t, flags=quote(error(fix)))
```

Finally, we can simply provide the correct solution to the problem wit the flag `ignore`. 

```{r, results="asis"}
# It prints the correct version
mathout(t, flags=quote(error(ignore)))
```


# Conclusions

This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them smoothly. Building on current features of R\ and existing packages for displaying formulas in R\ (e.g. Murrell and Ihaka [-@murrell2000], Allaire et al. [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML. Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax. 

The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.


In its current version \pkg{mathml} has some limitations. For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation. However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations. Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications. 

Also: no linebreaks in long equations. Mostly because it is not properly
supported by MathML renderers, and also in Latex, package "breqn" (insert
reference) is mostly a proof of concept. No support for `\begin{align*}` with
manual linebreaks etc.

\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`. The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.

## Mistakes

The case study presented above shows that \pkg{mathml} can help to improve
data analyses and statistical reports from an aesthetical perspective, as well
as regarding reproducibility of research. Furthermore, the package may also
allow for a better detection of possible mistakes in R programs.

Similar to most programming languages [@green1977], R\ code is notoriously hard
to read, and the poor legibility of the language is one of the main sources of
mistakes. For illustration, we consider again Equation 10 in
Schwarz [-@schwarz1994].

````{r, results="asis"}
f1 <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"]) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))))
}

mathout(f1)
````

The first version has a wrong parenthesis, which is barely visible in the code,
whereas the wrong curly brace in the mathematical representation is immediately
obvious. The correct version is shown below for comparison.

````{r, results="asis"}
f2 <- function(tau)
{ dfrac(c, mu["A"]) + (dfrac(1L, mu["A"]) - dfrac(1L, mu["A"] + mu["B"])) * 
    ((mu["A"]*tau - c) * pnorm(dfrac(c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau)))
      - (mu["A"]*tau + c) * exp(dfrac(2L*mu["A"]*tau, sigma["A"]^2L))
        * pnorm(dfrac(-c - mu["A"]*tau, sqrt(sigma["A"]^2L*tau))))
}

mathout(f2)
````

As the reader may know from own experience, missed parentheses are frequent
causes of wrong results and errors that are hard to locate in programming code.
This particular example shows that mathematical rendering can help to
substantially reduce the amount of careless errors in programming.

# Acknowledgments
 
## Appendix

Please check if the examples below

* are already presented somewhere
* can be added to one of the tables
* or if they should go into a separate appendix

_I removed the examples since they were already present._
