---
title: 'Report Schwarz (1994)'
author: "Irene Alfarone"
date: "11/7/2022"
bibliography: bibliography.bibtex
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Report Schwarz (1994)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A case study

As a case study, we present here the work of Schwarz [-@schwarz1994]. The reader
may recognize that this study can be traced back to the strand of
decision-making models, where decision-making is assumed to be a process of
noisy accumulation of information over time [e.g., @ratcliff2016]. The aim of
Schwarz' study is to present a new explanation of redundancy gains when
observers make speeded responses to stimuli of different sources, and the same
information is presented on two channels. Schwarz describes a model that assumes
the superposition of channel-specific diffusion processes that eventually reach
an absorbing barrier to elicit the response. For a detailed description the
reader may refer to the original article.


```{r, echo=FALSE}
library(rolog)
library(mathml)

# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

## The model 

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (redundant stimuli), with the redundant stimuli being
presented either simultaneously or with onset asynchrony \(\tau\). The channel
activation is described as a two-dimensional Wiener process with
drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
correlation \(\rho_\mathrm{AB}\) and initial
conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
process reaches the absorbing barrier \(c > 0\) for the first time. In combined
stimuli ("redundant" stimuli), the overall diffusion process
is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B
  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\). 

In single-target trials, the first passages at \(c\) are expected at
\(E\left[D_i\right] = \frac{c}{\mu_i}, i = A, B\), whereas in synchronous
redundant-target trials, we have
\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
stimulus onset asyncrony \(\tau > 0\) (Eq.\ 10),

````{r, results="asis", echo=FALSE}
f <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{ dfrac(c, mu_A) + (dfrac(1L, mu_A) - dfrac(1L, mu_A + mu_B)) * 
    ((mu_A*tau - c) * pnorm(dfrac(c - mu_A*tau, sqrt(sigma_A^2L*tau)))
      - (mu_A*tau + c) * exp(dfrac(2L*mu_A*c, sigma_A^2L))
        * pnorm(dfrac(-c - mu_A*tau, sqrt(sigma_A^2L*tau))))
}

mathml(f)
````

The observable response time is assumed to be the sum of \(D\), the time
employed to reach the threshold for the decision, and \(M\), denoting the of the
other processes.

\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)

## Methods

Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
a redundant signals task with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where 0 refers to synchronous AB, \(\infty\) refers to the single-target
presentation, and negative \(\tau\) denote those conditions in which B is
presented before A. Each condition was replicated 400 times. The 13 mean RTs and
standard deviations are given in Table 2 below.

Then, from Schwarz's (1994) Equations [2] [3] and [10], we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.34, \sigma_ \mathrm A = 4.3, \sigma_ \mathrm B = 11.7, c = 100, \mu_\mathrm M = 160\)


````{r, echo=FALSE}
mu_A = 0.53
mu_B = 1.34
sigma_A = 4.3
sigma_B = 11.7
tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
c = 100
mu_M = 160
````



```{r, results="asis", echo=FALSE}

# Conditions

g <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B)
{
    if( tau == Inf) return(dfrac(c, mu_A));
    if( tau == -Inf) return(dfrac(c, mu_B));
    if( tau > 0L)  return(f(tau, c, mu_A, sigma_A, mu_B, sigma_B));
    if( tau == 0L) return(dfrac(c, mu_A + mu_B));
    if( tau < 0L) return(f(-tau, c,  mu_B, sigma_B, mu_A, sigma_A))
}

mathml(g)

```


For instance, we can see that with \(\tau = -167\) the expected first passage time is:


```{r, results= "asis", echo=FALSE}
mu_A = 0.53
mu_B = 1.34
sigma_A = 4.3 
sigma_B = 11.7
tau = -167
c = 100
mu_M = 160

g(tau, c, mu_A, sigma_A, mu_B, sigma_B)

```


To which we have to add \(\mu_M = 160\) (which represents the amount of time that the movement of providing the answer takes) and we obtain: 


```{r, results= "asis", echo=FALSE}
g(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
```

Similarly, we obtain the first-time passage for every \(\tau\):



```{r, results= "asis", echo=FALSE}

G <- Vectorize(g, vectorize.args = 'tau')

h <- function(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  G(tau, c, mu_A, sigma_A, mu_B, sigma_B) + mu_M
}
mathml(h)
```



```{r, results= "asis", echo=FALSE}

mu_M = 160
mu_A = 0.53
mu_B = 1.34
sigma_A = 4.3 
sigma_B = 11.7
tau = c(-Inf, -167, -133, -100, -67, -33, 0, 33, 67, 100, 133, 167, Inf)
c = 100

# (h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M))
library(knitr)

```




### Model fitting and Parameters Estimation

Then we proceeded with the model fitting, using the data provided by Miller's (1986) study in Schwarz (1994). First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\). 

We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed (Ratcliff & McKoon, 2008) with at the first passage of c \(X(t) \sim N(\mu t , \sigma^2 t)\). Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\). Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)
  

For these reasons, in order to obtain the Goodness of Fit measure we proceeded with a z-standardisation.

```{r, results= "asis", echo=FALSE}


zf <-  function(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
{
  E <- h(tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
  dfrac(M-E, s/sqrt(N))
}

mathml(zf)

```




```{r, results= "asis", echo=FALSE}

# The vector X contains the observed values, the vector s their observed standard deviation and N = 400 are the number of trials. 

M = c(231, 234, 230, 227, 228, 221, 217, 238, 263, 277, 298, 316, 348)
s = c(56, 58, 40, 40, 32, 28, 28, 28, 26, 30, 32, 34, 92)
N = c(400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400, 400) 

zn <- zf(M, s, N, tau, c, mu_A, sigma_A, mu_B, sigma_B, mu_M)
# zn

```



Then, we calculated the Goodness of Fit as follows: 

\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).



```{r, results= "asis", echo=FALSE}

par <- c(mu_A = 0.53, sigma_A = 4.3, mu_B = 1.34, sigma_B = 11.7, mu_M = 160)

# Goodness of Fit

gof <- function(par, tau, M, s, N)
{
  zn <- zf(M, s, N, tau, c=100L, mu_A=par["mu_A"], sigma_A=par["sigma_A"], mu_B=par["mu_B"], sigma_B=par["sigma_B"], mu_M=par["mu_M"])
  sum(zn^2L)
}

mathml(gof)

n <- gof(par, tau, M, s, N)
# n

# pchisq(n, 8, lower.tail = FALSE, log.p = FALSE)

```



```{r, results= "asis", echo=FALSE}

lsq = function(tau, M, s, N)
{
  start = c(mu_A = 0.3, sigma_A = 2, mu_B = 1 , sigma_B = 5, mu_M = 200)
  optim(start, fn = gof,
    tau = tau, M = M, s = s, N = N, hessian = TRUE, control = list(fnscale = 1))
}

# mathml(lsq)

# Parameters' estimates
fit <-  lsq(tau, M, s, N)

est = fit$par
se = sqrt(diag(solve(fit$hessian)))
cilo = qnorm(0.025, est, se)
ciup = qnorm(0.975, est, se)

# Lower bound
#cilo

# Upper bound
#ciup

```

### Results

The calculated Goodness of Fit value is \(X^2= 29.43\), \(p= 0.0003\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.

The values of the estimated parameters reported in the table below are consistent with those reported by Schwarz (1994).

: Table 1: Estimated Model's Parameters

| Parameters | Estimate | 95% Confidence Interval
| :---: | :---: | :---:|
| \( \mu_ \mathrm A \)| 0.53 | 0.51, 0.56
| \( \sigma_ \mathrm A \)| 4.4 | 3.8, 5.1
| \( \mu_ \mathrm B \)| 1.3 | 1.2, 1.4
| \( \sigma_ \mathrm B \)| 18.3 | 11.6, 24.9
| \( \mu_ \mathrm M \)| 160 | 157, 165


Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i. \(\tau= 100\)) Miller's (1986) observed values are quite similar to the one predicted by Schwarz's (1994) model. Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz's (1994) sufficiently fit the data. 

: Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's (1989) study in Schwarz (1994)

| SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
|:--------:|:-----------------:|:----------------:|:-----------:|
\(-\infty\)| 235 | 231 | 56 |
\(-167\)   | 232 | 234 | 58 |
\(-133\)   | 231 | 230 | 40 |
\(-100\)   | 229 | 227 | 40 |
\(-67\)    | 227 | 228 | 32 |
\(-33\)    | 222 | 221 | 28 |
\(0\)      | 213 | 217 | 28 |
\(33\)     | 237 | 238 | 28 |
\(67\)     | 261 | 263 | 26 |
\(100\)    | 282 | 277 | 30 |
\(133\)    | 299 | 298 | 32 |
\(167\)    | 312 | 316 | 34 |
\(\infty\) | 349 | 348 | 92 | 
 _Note_.  Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.34, \sigma_ \mathrm A = 4.3, \sigma_ \mathrm B = 11.7, c = 100, \mu_\mathrm M = 160\) 
